Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-05 paper/paper.tex
restructured compilation section

Efficient implementation of expressive probabilistic programming languages has
recently been an active idea of research~\cite{MMR+07,GMR+08,P09,MWG+10,MSP14,
SDT14,WVM14,GS15,GSX+16}. There is often a compromise between the
expressiveness of the language and the efficiency of inference.  Some languages
emphasize expressiveness~\cite{MMR+07,GMR+08,MSP14,GS15}; others restrict the
class of models which can be expressed by the language~\cite{MWG+10,SDT14,NN14}
to facilitate application of efficient inference algorithms. Anglican allows
unrestricted specification of probabilistic models in the spirit of
Church~\cite{GMR+08}, while still supporting efficient scalable
inference~\cite{WVM14,PWD+14,RNL+2016}.

@@

Probabilistic programming languages are often implemented as pure functional
languages~\cite{GMR+08,MSP14,SGG15}. This is because reasoning about
probabilistic programsand distribution defined by the programs is easier in
pure functional setting. Anglican is also implemented as a language without
side-effects. However, there are also languages which follow imperative
paradigm and support mutable state directly~\cite{SDT14,GSX+16}, or through
probabilistic primitives~\cite{GMR+08,MSP14}. Anglican introduces a functional
alternative to stateful random primitives in the form of random processes.

@@

\subsection{Implementation Highlights}
\label{seq:highlights}

\iftoggle{full}{%
So far we introduced the basics of Anglican compilation to
Clojure. The described approaches and techniques are important
for grasping the language implementation but relatively well-known.
}{%
Anglican compilation to Clojure is based on CPS transformation
and compilation techniques, many of which are
well known and widely used~\cite{AJ89,A07}.
}%
In the rest of the section we focus on challenges we met and
resolved while implementing Anglican, as well as on
implementation of unique features of Anglican as a probabilistic
programming language.

\subsubsection{Continuations}
\label{seq:continuations}

@@

\subsubsection{Primitive Procedures}
\label{seq:primitive}

@@

\subsubsection{Probabilistic Forms}
\label{seq:forms}

In addition to checkpoints, there are a few other special forms
---  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
modify program state. \iftoggle{full}{These forms are translated into
expressions involving calls of functions from the
\texttt{anglican.state} namespace. }{}The \texttt{mem} form, which
implements memoization, deserves a more detailed explanation.

\paragraph{Memoization}
The author of a probabilistic model might want to 
randomly draw a feature value for each entity in a collection,
but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved
through the use of memoized functions. In Anglican, one
might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person]
                       (sample
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (eye-color 'bill)
    (eye-color 'john)))
\end{lstlisting}
The \texttt{mem} form converts a function to a memoized variant,
which remembers past inputs and the corresponding outputs, and 
returns the remembered output when given such a past input, 
instead of calling the original function with it. As a result,
for every input, random draws will be made only for the first time 
that the memoized function is called with the input; all subsequent
calls with the input will just reuse these draws and return the same output.

>>>>

Efficient implementation of expressive probabilistic programming languages has
recently been an active area of research~\cite{MMR+07,GMR+08,P09,MWG+10,MSP14,
SDT14,WVM14,GS15,GSX+16}. There is often a compromise between the
expressiveness of the language and the efficiency of inference.  Some languages
emphasize expressiveness~\cite{MMR+07,GMR+08,MSP14,GS15}; others restrict the
class of models which can be expressed by the language~\cite{MWG+10,SDT14,NN14}
to facilitate application of efficient inference algorithms. Anglican allows
unrestricted specification of probabilistic models in the spirit of
Church~\cite{GMR+08}, while still supporting efficient scalable
inference~\cite{WVM14,PWD+14,RNL+2016}.

@@

Probabilistic programming languages are often implemented as pure functional
languages~\cite{GMR+08,MSP14,SGG15}. This is because reasoning about
probabilistic programs and distributions defined by the programs, as well as
implementation of certain inference algorithms~\cite{MYM+15}, is easier in pure
functional setting. Anglican is also implemented as a language without
side-effects.  However, there are also languages which follow imperative
paradigm and support mutable state directly~\cite{SDT14,GSX+16}, or through
probabilistic primitives~\cite{GMR+08,MSP14}. Anglican introduces a functional
alternative to stateful random primitives in the form of random processes.

@@

\iftoggle{full}{%
\subsection{Implementation Highlights}
\label{seq:highlights}

So far we introduced the basics of Anglican compilation to
Clojure. The described approaches and techniques are important
for grasping the language implementation but relatively well-known.
In the rest of the section we focus on challenges we met and
resolved while implementing Anglican, as well as on
implementation of unique features of Anglican as a probabilistic
programming language.
}{%
}%

\iftoggle{full}{\subsubsection}{\subsection}{Probabilistic Forms}
\label{seq:forms}

\iftoggle{full}{\subsubsection }{\subsection}{Memoization}

\iftoggle{full}{%
In addition to checkpoints, there are a few other special forms
---  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
modify program state. These forms are translated into
expressions involving calls of functions from the
\texttt{anglican.state} namespace. The \texttt{mem} form, which
implements memoization, deserves a more detailed explanation.
}{%
}%

The author of a probabilistic model might want to 
randomly draw a feature value for each entity in a collection,
but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved
through the use of memoized functions. In Anglican, one
might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person]
                       (sample
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (eye-color 'bill)
    (eye-color 'john)))
\end{lstlisting}
The \texttt{mem} form converts a function to a memoized variant,
which remembers past inputs and the corresponding outputs, and 
returns the remembered output when given such a past input, 
instead of calling the original function with it. As a result,
for every input, random draws will be made only for the first time 
that the memoized function is called with the input; all subsequent
calls with the input will just reuse these draws and return the same output.

\iftoggle{full}{\subsubsection{Continuations}}{\subsection{Managing stack size}}
\label{seq:continuations}

@@

\iftoggle{full}{\subsubsection}{\subsection}{Primitive Procedures}
\label{seq:primitive}

@@
